\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-09-10 ResponseToReview.tex

\res{We agree entirely with this assessment of the Neighbor Joining trees. 
Our intention with the NJ trees was to show that there is no clear signal of selection on the \emph{Hopscotch} genotype across sequence in the \emph{tb1} genomic region.
Presumably, if the \emph{Hopscotch} element were under strong selection in teosinte populations we would expect to see \emph{tb1} haplotype groups in an NJ tree due to linkage disequilibrium and Hill-Robertson effects.
In fact, the original discovery of selection on the \emph{Hopscotch} insertion during maize domestication was based on such a signal from linked sequence 60kb distant from the \emph{Hopscotch} \citep{Studer2011}. We agree with the reviewer that the trees could be confusing and difficult to interpret, and the large polytomies do not provide information beyond our assertion that there is not strong haplotype structure broadly in the \emph{tb1} genomic region that correlates with the \emph{Hopscotch} genotype.
We have therefore opted to keep the NJ trees in the supplement where interested readers can view them but where they don't unnecessarily complicate our main story.
In addition, we have added explanation of the accession names in the supplemental figure legend.}

Phenotyping experiments - These are curious experiments.  First, why did you not generate a population of individuals with known genotypes and then test for the effect of genotype on phenotype?  Why take a random sample and just hope that you will get the necessary number of individuals of each genotype?  How can you randomize the growing conditions with respect to genotype, if you don't know genotypes ahead of time but instead determine them after all the growing measurements are taken?\\

\res{We agree that a planned grow-out with previously determined genotypes could be useful.  
However, prior work has already shown that \emph{Hopscotch} has a phenotypic effect in isogenic and inbred backgrounds \citep{Studer2011}, and our interests lie in the effects of the \emph{Hopscotch} in genetic backgrounds occurring in nature.
Moreover, our sampling of individuals from this population for our grow-out was not random.  
We initially genotyped a single individual from each sampling site within the population and then targeted our sample to include individuals from sampling sites where individuals were homozygous or heterozygous for the \emph{Hopscotch} allele.
Our reasoning in using this approach was that we wanted a high proportion of both \emph{Hopscotch} positive and \emph{Hopscotch} negative individuals for phenotypic comparisons and, given that the \emph{Hopscotch} allele is typically rare in teosinte, we wanted to enrich our sample for this allele. 
We were overly successful in this approach and had a higher frequency of individuals with the \emph{Hopscotch} positive allele.
We have clarified our sampling methodology on page 9, lines 188-206. Finally, plants were situated randomly in the greenhouse prior to genotyping, which effectively represents a randomized experimental design.}

@@

Why are sample sizes not reported?  You provide total samples sizes (eg, 206 or 216 for Phenotyping 1, depending on whether you read the Methods or the Results), but not sample sizes for each genotypic class.\\

\res{We have reported sample size information for all three genotypes on page 12 lines 281-284.}

@@

They present an interesting result, their conclusions are supported by their data, and the methods they use are appropriate. However, I feel like they need to make some further effort to understand what is happening mechanistically. What is Hopscotch doing in maize vs these sampled teosintes? I feel like the expression analyses and phenotyping experiments they suggest should be part of a later study should be part of this study. At a minimum, qRT-PCR experiments should be performed assessing tb1 expression in their with/without Hopscotch lines. Expression analyses of gt1, te1, and tru1 would be great, since they invoke variation at these loci to explain their results.\\

\res{Extensive expression analysis of individuals with and without the \emph{Hopscotch} insertion at \emph{tb1} and a number of other genes is a good suggestion for follow-up work but would be a substantial undertaking and indeed more appropriate for another paper. }

Another potential avenue, that the authors also suggest, lies in more directed and exhaustive greenhouse experiments. Variation in red and far-red light has been shown to be important in regulating both tb1 (in Sorghum) and gt1 in maize. Growth chamber experiments (with or without supplemental far red light, with or without Hopscotch), coupled to expression analyses in seedlings, seem like they should be possible.  In the greenhouse experiments that were done, phenotyping was incomplete. Apart from tillering, Hopscotch and natural variation at the tb1 locus appears to affect ear architecture and internode length. Was there any evidence for variation in these other traits? In the longer term, it would be nice to see what happens when you introgress these particular parviglumis loci into maize.\\

\res{We did not measure ear architecture or internode length. 
While \emph{tb1} does impact these traits, our primary interest was in the effect of the \emph{Hopscotch} insertion at \emph{tb1} on tillering given that this is an ecologically relevant phenotype that is known to be substantially affected by \emph{Hopscotch} in a maize background.
Both additional greenhouse and expression analyses would indeed be useful avenues of further investigation.
Nonetheless, these are extensive additional experiments which we are unable to include at this time. 
We have, however, added text to the conclusion (Page 15, lines 389-393) of this paper discussing their utility and the need for further work to elucidate the role of \emph{tb1} in branching architecture in teosinte:
``Future studies should examine expression levels of \emph{tb1} in teosinte with and without the \emph{Hopscotch} insertion and further characterize the effects of additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{te1}). These data, in conjunction with more exhaustive phenotyping, should help to further clarify the ecological significance of the domesticated \emph{tb1} allele in natural populations of teosinte."}

@@

One minor issue: in Fig. 1, it would be nice to see where the Balsas River Basin is in relation to the sampling shown.\\

\res{We have added the Balsas River in a revised version of Figure 1.}

\section*{Response to Reviewer 2} 
5) Although this paper brings up an interesting question regarding the role of tb1 standing variation in teosinte, it is not able to answer it completely.  It presents results that have been published previously about the presence of hopscotch in teosinte, maize, and mexicana.\\

\res{Reviewer 2 is correct in stating that it was known that the Hopscotch was present in some teosinte individuals \citep{Studer2011}; however, previous studies sampled a small fraction of the populations and individuals per population that we present here and thus did not provide comprehensive estimates. 
Our study is not only concerned with the presence or absence of the \emph{Hopscotch} allele in teosinte but also the frequency at which it segregates in a range-wide sample of teosinte.  
This finer level of detail sheds light on the evolutionary history of the \emph{Hopscotch} insertion (\emph{i.e.}, the domestication allele at \emph{tb1}) in teosinte. 
We have added clarifying text to the introduction (Page 5, lines 59-64): ``TThe effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes \citep{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte. Previous studies have confirmed the presence of the \emph{Hopscotch} in samples of \emph{parviglumis}, \emph{mexicana}, and landrace maize; however little is known about the frequency with which the \emph{Hopscotch} is segregating in natural populations."}

@@

Nevertheless, the manuscript is worth publishing as it does present data that may help rule out some possible explanations as to why tb1 is so common in parviglumus.\\

\res{Thank you; we hope that our paper in conjunction with future research and experiments can help shed light on the ecological significance of both \emph{tb1} and the \emph{Hopscotch} insertion in teosinte.}

6) This paper is fairly maize-centric. A more general message is included in the discussion, not as much in the introduction, and not at all in the abstract. This could be easily remedied.\\

\res{We have editied the introduction to include a more broad explanation of domestication and shade avoidance responses in plants. Edits can be seen on page 4, lines 24-38 and page 5, lines 65-70.}

@@

Methods and statistics were appropriately used as far as my expertise allows me to judge.
Specific suggestions follow:

7) Introduction:
Page 4, Lines 7 - 9: You say ``Additionally, many domesticates show reduced genetic diversity when compared to their wild progenitors, and an understanding of the distribution of diversity in the wild and its phenotypic effects has become increasingly useful to crop improvement \citep{KovachMcCouch2008}.".  Can you tell us why an understanding is useful? Can you give some examples?\\

\res{We have expanded on this concept in the introduction on page 4, lines 24-40: ``Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. 
Plant domestication usually involves a suite of phenotypic changes such as loss of seed shattering and increased fruit or grain size, which are commonly referred to as the `domestication syndrome' \citep{OlsenWendel2013}, and much of the study of domestication has focused on understanding the genetic variation underlying these traits \citep{OlsenGross2010}. 
Because most domesticates show reduced genetic diversity relative to their wild counterparts, effort has been made to identify agronomically useful variation in crop wild relatives \citep{flint2009wide}. Often, after identification, the alleles conferring these beneficial traits are bred into domesticates for crop improvement. For example, \emph{Oryza rufipogon}, the wild progenitor of domesticated rice, has proven useful for the integration of a number of beneficial QTL controlling traits such as grain size and yield into domesticated rice \citep{KovachMcCouch2008}. In addition to researching the role of wild alleles in domesticates, researchers have also investigated the role of variation in domesticated taxa in the evolution of feral and weedy populations \citep{ellstrand2010crops}. But even though domesticated alleles are often found segregating in wild relatives \citep{Gallavotti2004, SigmonVollbrecht2010}, we know almost nothing about the ecological role of this variation in natural populations. In this paper we present an ecological genetic analysis of the domestication locus \emph{tb1}, and specifically the domesticated haplotype at \emph{tb1}, in natural populations of the wild ancestor of domesticated maize."}

@@

8) Page 5, line 12 - 14: You say "The tb1 locus appears to play an important role in the shade avoidance pathway in Zea mays and other grasses and may therefore be crucial to the ecology of teosinte (Kebrom and Brutnell, 2007; Lukens and Doebley, 1999)." Please clarify; do the references belong to the first half of the sentence, supporting tb1 playing a role in Zea mays, or the second half of the sentence, supporting tb1 possibly being crucial to the ecology of teosinte? Which of the two distinct things have actually been studied? If the first, could you tell us how tb1 has been shown to work in maize under shade conditions? If the second, was it actually studied, or merely hypothesized in teosinte?\\

\res{We have moved the references cited here to more clearly reflect the papers' findings.  We have also added text describing the role of \emph{tb1} in the shade avoidance pathway and clarifying the manner in which this could affect fitness in teosinte. Changes can be seen on page 5, lines 65-75: ``In teosinte and other plants that grow at high population density, individuals detect competition from neighbors via the ratio of red to far-red light.  An increase in far-red relative to red light accompanies shading and triggers the shade avoidance syndrome, a suite of physiological and morphological changes such as reduced tillering, increased plant height and early flowering \citep{KebromBrutnell2007}. The \emph{tb1} locus appears to play an important role in the shade avoidance pathway in \emph{Zea mays} \citep{LukensDoebley1999} and other grasses \citep{KebromBrutnell2007} via changes in expression levels in response to shading. \citet{LukensDoebley1999} introgressed the teosinte \emph{tb1} allele into a maize inbred background and noted that under low density conditions plants were highly tillered but that under high density, plants showed significantly reduced tillers and grew taller. 
Based on these results we hypothesize that the \emph{Hopscotch} (\emph{i.e.}, the domesticated allele) at \emph{tb1} may play a role in the ecology of teosinte, especially in high-density populations.}

@@

9) Page 5, Lines 15 - 16: you say you ``aim to characterize the distribution of the Hopscotch insertion in parviglumis, mexicana, and landrace maize, and to examine the phenotypic effects of the insertion in parviglumis."  The mexicana seem a bit tacked on and not much discussed. Did you phenotype mexicana? Can you present the data here, if so? Does the analysis of this group of teosintes add anything to your conclusions?\\

\res{Our goal in screening for the \emph{Hopscotch} insertion was to assess its taxonomic and geographic distribution to the extent possible given the samples to which we had access.  Reviewer 2 is correct in noting that ssp. \emph{mexicana} is not as prevalent in our paper as ssp. \emph{parviglumis}.  Though our sample of \emph{mexicana} is range-wide, we only had access to seeds from 17 populations.  Frequency of the \emph{Hopscotch} insertion was low in these populations relative to \emph{parviglumis}, an interesting point that we have discussed further on page 13, lines 310-313 of the text.  The low frequency of \emph{Hopscotch} in \emph{mexicana} precluded a phenotyping experiment with this subspecies since a very high number of seeds would have to be germinated in order to obtain enough individuals with the \emph{Hopscotch} allele.}

@@

10) Materials and Methods:
Page 5, lines 22 - 23: how many individuals per accession? A fixed number, or variable, and depending on what? This can be mentioned so the reader does not have to go to the appendix or supplements.\\

\res{We have added a range of values (1-38 individuals) per population in the methods, pages 5-6, lines 83-88: ``We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{mexicana} populations, and 86 \emph{parviglumis} populations; ranging from 1-38 individuals per population, with an average of 11 individuals per population for \emph{parviglumis} and \emph{mexicana} and 2 individuals per landrace accession) and assessed the presence or absence of the \emph{Hopscotch} insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article)."}

@@

11) Page 5, Line 28: mention how big the hopscotch element is.  Can it be amplified in one amplicon? Again, this would be useful to know without having to look in appendix or supplements.\\

\res{The sizes of amplicons are now mentioned explicitly in the Materials and Methods (page 5, lines 97-102) following primer sequences: 
``Homozygotes show a single band for absence of the \emph{Hopscotch} element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes are three-banded (Appendix 2, See Supplemental Materials with the online version of this article)."
The 5kb amplicon including the entire \emph{Hopscotch} could be amplified in a single reaction.}

@@

12) Page 6 line 6: "When only one PCR resolved well, we scored one allele for the individual".  Sorry, what does this mean? That you have one band (one assumes the short one?). If you are talking presence/absence, how do you know the allele?\\

\res{We developed primers for two PCR reactions that would amplify products with size polymorphism characteristic of the two different alleles (\emph{i.e.}, \emph{Hopscotch} vs. no \emph{Hopscotch}). If PCR products were clearly resolved for only one of these reactions, we included one allele for that individual in our data set rather than infer the diploid genotype. We have clarified this in the text on page 6, lines 92-102. : ``Two PCRs were performed for each individual, one with primers flanking the \emph{Hopscotch} (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
HopR, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, {\small  5'-CCTCCACCTCTCATGAGATCC-3'} (Appendix 3 and Appendix 4, See Supplemental Materials with the online version of this article). Homozygotes show a single band for absence of the element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes show all three bands (Appendix 2, See Supplemental Materials with the online version of this article). Since we developed a PCR protocol for each allele, if only one PCR resolved well, we scored one allele for that individual rather than infer the diploid genotype."}

13) Page 7 lines 14 - 15 "These analyses only included populations in which eight or more individuals were sampled."  How many populations did that include?\\

\res{Thirty-two populations were included in these analyses, and we have edited the text to reflect this number on page 6, lines 108-109: ``These analyses only included populations ($n = 32$) in which eight or more chromosomes were sampled."}

@@

14) Page 7 lines 17 - 26: how many year's worth of environmental data went into your analysis? Can you hypothesize that tb1 differences can be correlated with current climactic data if the changes and subsequent selections must have happened many generations ago?\\

\res{Our climatic data consisted of the 19 bioclim layers commonly used in ecological niche analyses. Full methods for generation of these data can be found at the following url: \url{http://www.worldclim.org/methods}. Briefly, temperature and precipitation were averaged for each global location over a minimum of ten years during the period from 1950-2000.  Soil data were downloaded from the Harmonized World Soil Database (FAOHWSD; \url{http://webarchive.iiasa.ac.at/Research/LUC/External-World-soil-database/HWSD_Documentation.pdf}). This recently published data set drew together global soil maps generated by a world-wide network of scientists over the last several decades and summarized them in GIS data layers that are available as polygon coverages. Data for our sampling sites were extracted from these data sets based on latitude and longitude using the DIVA-GIS platform. Since data for our analyses were either summarized over the last several decades (climatic data) or are likely to be stable over time (soil data), they have potentially influenced current frequencies of the \emph{Hopscotch} insertion in our sampled populations. We have further clarified this on page 7, lines 119-124 of the text : ``Environmental data were obtained from \url{www.worldclim.org} and soil data were downloaded from the Harmonized World Soil Database \citep{FAOHWSD} at \url{www.harvestchoice.org}.  Environmental data represent average values for the last several decades (climatic data) or are likely stable over time (soil data) and therefore represent conditions important for local adaptation of our samples.  Information from these data sets was summarized by principle component analysis following \citet{Pyhajarvi2013}."}

@@

15) Page 7 lines 22 - 23: "We used genotyping and covariance data from Pyhajarvi et al. (2013) for BayEnv" what kind of dataset is this? Teosinte?\\

\res{We have clarified the dataset in the paper on pages 6-7, lines 116-119: ``We used teosinte (ssp. \emph{parviglumis} and ssp. \emph{mexicana}) genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the \emph{Hopscotch} insertion coded as an additional biallelic marker. SNP data from \citep{Pyhajarvi2013} were obtained using the MaizeSNP50 BeadChip and Infinium HD Assay (Illumina, San Diego, CA, USA)."}

@@

16) Page 8 lines 20 - 21:" We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes." Does LD data in your parviglumus support 5Kb?\\

### 2014-09-10 ResponseToReview.tex

\res{While the recombination landscape will vary to some extent across individuals of teosinte and maize, broad-scale patterns are conserved \citep{bauer2013}.  We therefore relied on previous estimates of the decay of linkage disequilibrium based on data from HapMap2 of maize that suggest LD decays over an average distance of 5500bp \citep{Chia2012}; because LD decay is even more rapid in teosinte \citep{Chia2012} we used a conservative haplotype block size of 5kb as an estimate of the extent of linkage disequilibrium in our sample and the physical distance over which SNPs would have to be separated in order to be in weak linkage and not violate assumptions of the STRUCTURE model.}

17) Page 9 lines 2 - 3: ``Plants were watered three times a day by hand and with an automatic drip containing 10-20-10 fertilizer." This sounds odd; by hand or automatically? Or you turned on the drip by hand? In which case, it was not automatic\\

\res{We have edited the text to clarify, page 9, lines 198-200 : ``Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer, which was supplemented with hand watering on extremely hot and dry days."}

@@

18) Page 9 lines 7 - 8: ``Culm diameter is not believed to be correlated with tillering index or variation at tb1." So, why did you do it? Because of the last sentence in the materials and methods? Perhaps do not mention here that is not correlated; wait until the end of the materials and methods where you seem to explain it.\\

\res{Thank you for the suggestion, we have edited the text accordingly and have added our explanation of the inclusion of culm diameter at the end of the Materials and Methods section, pages 9-10, lines 212-216: ``Additionally, in order to compare any association between \emph{Hopscotch} genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS. Culm diameter is not believed to be correlated with tillering index or variation at \emph{tb1} and is used as our independent trait for phenotyping analyses."}

@@

19) Page 9 lines 9 - 10: "We genotyped individuals for the Hopscotch insertion following the protocols listed above." The PCR protocols or the sequencing protocols?\\

\res{We have clarified in the manuscript that genotypes were obtained using our PCR protocol.}

20) Results:
Page 10 lines 5 - 6: ``Within our parviglumis and mexicana samples we found the Hopscotch insertion segregating in 37 and four populations, respectively"; remind us how many parviglumis and Mexicana samples you had total? So we can see if this comes out to a very different ratio than in maize."\\

\res{We have added ratios to the text, page 10, lines 222-225: ``Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico."}

### 2014-10-24 AJB_Res2.tex
Edits to AJB Res 2
Comments in the response as to what can be addressed.

@@

@@

The main problem for me about this argument is that there is an implicit link between the presence or absence of Hopscotch and levels of tb1 expression in teosinte plants, and the effect this has on tillering.  There is excellent data in this manuscript that deserves to be published in some form, but I do not believe that the argument that is constructed is valid without showing that Hopscotch actually affects tb1 expression in teosinte. It would be a different matter if the paper were merely focused on the effect of the presence or absence of Hopscotch, but the paper is written explicitly assuming a connection between Hopscotch and tb1 expression in teosinte, and often conflates variation at the tb1 locus with presence or absence of Hopscotch.  This is a problem, especially when the data show little effect of Hopscotch presence or absence on the measured phenotypes.  Below I expand on this summary and also present some other queries.

Introduction
page 5, lines 21-28.  The logic here is unclear.  The Lukens and Doebley experiment showed that maize lines with an introgressed teosinte tb1 allele were affected by shading.  It does not follow that the domesticated allele (Hopscotch containing allele) should play a role in teosinte ecology, especially since maize with the Hopscotch allele is insensitive to shading.  This argument would make more sense if the authors had shown that Hopscotch alleles affect tb1 expression in teosinte.

@@

Materials and Methods
Page 6, line 30: Minor point, and I am not an expert on population genetics, but why say that you only include populations in which 8 or more chromosomes were sampled, rather than 8 or more individuals.  If I am missing something, it could be that others will too, so please explain.

Page 6, line 30- page 7, line 2: Testing the adaptive hypothesis.  I understand that in the end there was no evidence for Hopscotch allele frequency being correlated with environmental variables.  Was there evidence for any of the other SNPs being so correlated?  How robust is this analysis against false positives?  The adaptive hypothesis is attractive, but it is unclear why you would make it.  You do not make the claim in these lines, but it seems to me that it is the three classes of teosinte alleles that might be more likely to correlate with environment (since Studer et al. 2012 showed that they correlated with taxonomy) than the presence or absence of the Hopscotch element.

Page 9, line 15: In the STRUCTURE analysis, why was K=2 chosen?  Had you already done exploratory runs with a range of K values or is there some theoretical reason for assuming K=2.  Either way, please explain.

Page 10, line 12: Why is culm diameter not believed to correlated with tillering index?  Is there a reference for this? What do your results show?  More importantly, if you want to prove that Hopscotch has more effect on tillering than on other independent traits you really need to do more comprehensive phenotyping.  Please explain more fully what this particular analysis is meant to achieve.
["Culm diameter is not believed to be correlated with tillering index or variation at tb1 and is used as our independent trait for phenotyping analyses."]

Discussion
Page 14, 3rd para: This paragraph starts off by examining explanations of genetic drift and natural selection for different frequencies of Hopscotch alleles amongst teosinte populations.  After discussing evidence for bottlenecks, the argument shifts to the lack of diversity in the 5' UTR of the tb1 gene, concluding that the results are consistent with the action of selection on the 5'UTR, and that there might be an ecological role for the gene in some populations.  This is a classic example of the veering in focus between the purported object of this study, hopscotch elements, and the tb1 gene itself.  In the end there is no explanation given for the hopscotch frequencies, instead the focus is switched back to the gene itself.  To the reader there is an implication (perhaps unintentional) that there is a causative link between Hopscotch and variation in the tb1 alleles, whereas I would say this has simply not been shown in teosinte, and the evidence in this manuscript would
argue against such a link.

Page 15, line 10: the argument at this point appears to be that we don't see the effect of Hopscotch in teosinte because of variation in other loci.  I would like more explanation of how that might occur, given that the Hopscotch element in maize upregulates tb1 expression so much that it explains most of the variation.

Page 15, line 14, discussion of Studer et al. (2012).  Studer et al. showed that there is an allelic series of tb1 in teosinte, and that this series is correlated with taxonomy of maize subspecies.  They also showed that differences between the three classes of tb1 alleles were significantly correlated with mean internode length of the uppermost lateral branch as well as two inflorescence traits, but that the classes did NOT correlate with variation in tillering.  Studer et al. concluded that there was evidence for allelic series only for the internode length and inflorescence characters, not for tillering.  Their explanation was NOT that tillering index was controlled by both tb1 and other loci, but that there was no evidence for variation in tb1 alleles affecting tillering.  Changes in other loci were invoked by Studer et al. (along with environmental effects on seed development) to explain why separate families differed from each other irrespective of the tb1 allele they
contained.  Again, at the end of this paragraph (page 15, lines 27-28), there is confusion over what is being explained, as the last sentence refers to a lack of correlation between Hopscotch genotype and tillering, but the discussion of the Studer et al. (2012) paper does not concern Hopscotch, but only an allelic series at the tb1 locus.  In fact, the Studer paper does not mention Hopscotch elements at all and the allelic series analyzed are only the coding region of the tb1 gene.

@@

>>>>

\lev{No on expression data and additional experiments. I thought our hypotheses/aims were rather clearly stated, but perhaps we can reword if you guys think that they need clarification.}

@@

\lev{Will check numbers. I think the confusion is in the individuals vs. alleles.}

@@

The main problem for me about this argument is that there is an implicit link between the presence or absence of Hopscotch and levels of tb1 expression in teosinte plants, and the effect this has on tillering.  There is excellent data in this manuscript that deserves to be published in some form, but I do not believe that the argument that is constructed is valid without showing that Hopscotch actually affects tb1 expression in teosinte. It would be a different matter if the paper were merely focused on the effect of the presence or absence of Hopscotch, but the paper is written explicitly assuming a connection between Hopscotch and tb1 expression in teosinte, and often conflates variation at the tb1 locus with presence or absence of Hopscotch.  This is a problem, especially when the data show little effect of Hopscotch presence or absence on the measured phenotypes.  Below I expand on this summary and also present some other queries.  
\lev{no expression analyses. should we downplay any mention of expression in the paper? or somehow make it more clear that it will be a nice follow up in another study? If all the reviewers we have had so far feel expression is crucial, I do think that we'll have to find a way to address the lack of expression in the paper}
Introduction
page 5, lines 21-28.  The logic here is unclear.  The Lukens and Doebley experiment showed that maize lines with an introgressed teosinte tb1 allele were affected by shading.  It does not follow that the domesticated allele (Hopscotch containing allele) should play a role in teosinte ecology, especially since maize with the Hopscotch allele is insensitive to shading.  This argument would make more sense if the authors had shown that Hopscotch alleles affect tb1 expression in teosinte. \lev{I don't understand. If Hopscotch = taller plant with fewer tillers then why couldn't it play a role in the ecology of teosinte? Our phenotyping obviously doesn't support this pattern in teosinte, but I don't understand how the hypothesis is a problem.}

@@

\lev{I don't think there is a big problem with this. I think it's clearly stated. Is the problem he is pointing out that we have explicit hypotheses about the phenotypic effects and that it may play a role in the ecology that we don't include in this statement?}
Materials and Methods
Page 6, line 30: Minor point, and I am not an expert on population genetics, but why say that you only include populations in which 8 or more chromosomes were sampled, rather than 8 or more individuals.  If I am missing something, it could be that others will too, so please explain.
\lev{This was just our sampling cut-off for analyses...}

Page 6, line 30- page 7, line 2: Testing the adaptive hypothesis.  I understand that in the end there was no evidence for Hopscotch allele frequency being correlated with environmental variables.  Was there evidence for any of the other SNPs being so correlated?  How robust is this analysis against false positives?  The adaptive hypothesis is attractive, but it is unclear why you would make it.  You do not make the claim in these lines, but it seems to me that it is the three classes of teosinte alleles that might be more likely to correlate with environment (since Studer et al. 2012 showed that they correlated with taxonomy) than the presence or absence of the Hopscotch element.  
\lev{I only used the Hopscotch, coded as a SNP. Tanja's paper would have included the other SNPs in the panel. I don't remember off the top of my head if she found any that were significant on Chromosome 1, but that is easily checked. Should we mention SNPs she identified in her paper her? Yes, the three classes of teosinte likely correlate with environment, but that's not something we address here. Should we take out BayEnv stuff all together? I like it in, but it doesn't add much I guess.}

Page 9, line 15: In the STRUCTURE analysis, why was K=2 chosen?  Had you already done exploratory runs with a range of K values or is there some theoretical reason for assuming K=2.  Either way, please explain.
\lev{I think we chose K=2 because we expected things to group either into a teosinte cluster or a maize cluster}

Page 10, line 12: Why is culm diameter not believed to correlated with tillering index?  Is there a reference for this? What do your results show?  More importantly, if you want to prove that Hopscotch has more effect on tillering than on other independent traits you really need to do more comprehensive phenotyping.  Please explain more fully what this particular analysis is meant to achieve.
["Culm diameter is not believed to be correlated with tillering index or variation at tb1 and is used as our independent trait for phenotyping analyses."]
\lev{I think we'd cite studer for this. we wanted a trait that would be independent so that we could compare genotype + culm patterns to genotype + tillering patterns.}
Discussion
Page 14, 3rd para: This paragraph starts off by examining explanations of genetic drift and natural selection for different frequencies of Hopscotch alleles amongst teosinte populations.  After discussing evidence for bottlenecks, the argument shifts to the lack of diversity in the 5' UTR of the tb1 gene, concluding that the results are consistent with the action of selection on the 5'UTR, and that there might be an ecological role for the gene in some populations.  This is a classic example of the veering in focus between the purported object of this study, hopscotch elements, and the tb1 gene itself.  In the end there is no explanation given for the hopscotch frequencies, instead the focus is switched back to the gene itself.  To the reader there is an implication (perhaps unintentional) that there is a causative link between Hopscotch and variation in the tb1 alleles, whereas I would say this has simply not been shown in teosinte, and the evidence in this manuscript would
argue against such a link.
\lev{I don't mind this paragraph. It ties in findings from previous studies that identified evidence of bottleneck in some of these pops with our findings, which i think is useful. We state that a population bottleneck could explain the pattern that we see in SLO, but that it cannot account for this pattern across multiple populations in the Jalisco region.}

Page 15, line 10: the argument at this point appears to be that we don't see the effect of Hopscotch in teosinte because of variation in other loci.  I would like more explanation of how that might occur, given that the Hopscotch element in maize upregulates tb1 expression so much that it explains most of the variation.  
\lev{There were other loci in the initial QTL paper that are involved in the difference in branching architecture. Plus we now know of a lot of other loci (mentioned in paper) that act in the SAS and could affect plant architecture.}

Page 15, line 14, discussion of Studer et al. (2012).  Studer et al. showed that there is an allelic series of tb1 in teosinte, and that this series is correlated with taxonomy of maize subspecies.  They also showed that differences between the three classes of tb1 alleles were significantly correlated with mean internode length of the uppermost lateral branch as well as two inflorescence traits, but that the classes did NOT correlate with variation in tillering.  Studer et al. concluded that there was evidence for allelic series only for the internode length and inflorescence characters, not for tillering.  Their explanation was NOT that tillering index was controlled by both tb1 and other loci, but that there was no evidence for variation in tb1 alleles affecting tillering.  Changes in other loci were invoked by Studer et al. (along with environmental effects on seed development) to explain why separate families differed from each other irrespective of the tb1 allele they
contained.  Again, at the end of this paragraph (page 15, lines 27-28), there is confusion over what is being explained, as the last sentence refers to a lack of correlation between Hopscotch genotype and tillering, but the discussion of the Studer et al. (2012) paper does not concern Hopscotch, but only an allelic series at the tb1 locus.  In fact, the Studer paper does not mention Hopscotch elements at all and the allelic series analyzed are only the coding region of the tb1 gene.
\lev{I think we should reread this section of the studer paper and check. If I remember correctly they found differences between their maize isogenic lines that were comparable to the differences they found among maize with introgressed segments of tb1}

@@

\lev{nothing to change for reviewer 2}
